An Event-based Architecture for Multi-population Optimization Algorithms

### 2020-01-27 draft.tex
Experimental Setup

\subsection{Event Driven Model} 
\label{edm}
Based on a reactive architecture we proposed in a previous work
\cite{guervos2018introducing}, we now describe the general architecture shown in
Figure \ref{fig:kafkEO}. Again, we can explain the model using the analogy
of Producers and Consumers of messages. First, we can see that
there are two queues, one labeled Input, and the other one Output. In the diagram, 
push operations on a queue are represented by solid arrows connecting to the left side
of the box, and pull or pop operations as solid arrows leaving from the right side.
Also, the architecture has at least four processes indicated in the diagram as
swimlanes: First, the Setup process, responsible for the reading a configuration
file, and creating the initial subpopulations. Second, a Controller process,
responsible for the migration between subpopulations, and keeping track of the
iterations of the algorithm. The Message Queue process runs the Input and Output
queues. Finally, there is at least one Stateless Function process responsible
for running the isolated algorithms. In the example, two processes PSO and GA,
are shown.

The algorithm starts with the Setup process pulling a configuration message from
the experiments queue, this is not shown in the diagram. The configuration message 
includes all the parameters needed to execute the algorithm, among other things,
the number of subpopulations, the number of individuals, and the number of iterations
of the algorithm. We give more details about the configuration data structure
in Section \ref{experiment_flow}. % Missing  % added 
Once the configuration is read, we can follow the path of 
messagess as follows:

\begin{enumerate}
\item In this step, the specified number of subpopulations is created according to the 
parameters found in the configuration structure. The population at this moment is just
static data, including each individual inside. Each subpopulation includes a metadata
section where its algorithm and execution parameters are specified. For instance, 
for a GA, the mutation rate, type of crossover operator, and other values are indicated.

\item Each subpopulation is then pushed to the Input queue, so they can be consumed 
by Stateless functions responsible for the execution of the search.  % Why the caps in Stateless? - JJ  
% It refers to the swimlines in the diagram, maybe I should point it out  - Mario
% Added a comment above. - Mario

\item One or more Stateless functions are constantly pulling subpopulation messages
from the Input queue. Taking the population data and metadata as parameters, these 
functions are responsible for the actual execution of the optimization algorithm. 
They take the current state of the population and run the algorithm for a certain 
number of iterations.

@@

\item The Controller process is responsible for keeping track of the progress of 
the search. It pulls current subpopulations from the Output queue, inspects the metadata 
and if an optimal solution has been found or the maximum number of iteration has been
reached it signals the stop of the execution.

\item Otherwise, it passes the stream of messages 
to a migration process, where subpopulations are mixed with each other. New populations 
are generated from this migration, and they are again pushed to the Input queue to continue 
in a loop. The Controller is also responsible of logging the metadata recived with the
messages.  
\end{enumerate}

@@

On the other hand, there are several caveats as well:
It is more costly to move entire subpopulations as messages than only passing certain individuals 
from a process to another. Designers must consider this cost when working with large
individuals, and if possible send sub-populations using several messages, use compression or 
adjust the size of subpopulations. Pool-based algorithms also suffer from this drawback, but 
web based implementations have been working with continuous optimization problems of 1000 dimensions.
But, in other cases this is not a viable approach.

@@

Next, we compare the message-driven architecture we presented against four other
proposals for multi-population algorithms found in the literature. We center the
comparison on the coupling and communication between the main components:
Subpopulations, Processing Nodes, and Algorithms.  In Figure \ref{fig:population_message} we show the
main components of a message-driven architecture. We can see that the algorithm
and subpopulations are separated, and the Population Queue keeps the state
of subpopulations. While algorithms only need the subpopulations as parameters
for their execution.

@@

In this section, we propose a design of a reactive container-based application
for executing multi-population based optimization experiments.  For the design,
we followed the design patterns highlighted in the previous section, and again
we can go back to Figure \ref{fig:kafkEO} and use it as a guide for explaining
the main components and their interactions. But first, we give a brief
description of the technology we used.

\subsection{Containers} 
\label{containers}
% TO DO

\subsection{Docker Compose} 
\label{compose}
% TO DO

% Do we need a description of every container? Wouldn't a general
% description of the containers used be enough? - JJ 
% Ok, I think only workers and the controller need to be described in detail - Mario

@@

The controller is an essential component of the architecture (Figure \ref{fig:kafkEO})
because it is responsible for maintaining the evolutionary loop. It takes newly
evolved populations from the Output Queue, mixes them,   and produces new
messages from the result. At the same time, it must keep track of two conditions
for ending the loop: the number of messages it has received reaches a maximum
value, or the error of the best solution goes below a specific threshold.  Finally, it
must filter out remnant messages from other experiments. Messages from other
experiments can remain in the queues because of the asynchronous nature of the
system.  We chose to implement the controller in Python but using an API
specialized in asynchronous event-driven programming over data streams.  
The open-source library is called Reactive Extensions (ReactiveX) for Python 
\url{https://github.com/ReactiveX/RxPY} an it is based on the Observer 
pattern \cite{gamma1995design} and functional programming.

@@

In ReactiveX, an observer object subscribes to an Observable. Observables emit a
sequence of items, and all the subscribed observers react to each emission. The
ReactiveX library includes several reactive operators that we can use to
transform and combine sequences of items. These operators provide reactive
extensions that allow us to compose asynchronous sequences together in a
declarative manner. In Figure \ref{fig:marble_controller}, we use a so-called marble diagram to
represent the composition of Observables and reactive extensions operators. We
represent the timeline of an Observable as a horizontal arrow, which indicates
that time flows from left to right. In the diagram, items are represented as
marbles. The position of each marble indicates the point in time when they were
emitted by the Observable. Reactive operators are represented as text boxes,
showing the transformation to be applied. Normally, a transformation results in
another Observable, again emitting new results.  For asynchronous programming, 
this approach is more elegant than nested callbacks, that are more difficult 
to code and debug.

@@

\begin{enumerate}

\item The controller continually pulls new messages from the Output 
message queue. These messages are instantly emitted by the  \texttt{consumed\_messages}
Observable. In this particular period, we can see that the stream receives 
two populations from another experiment, and are shown as two red marbles.  
The filter operator removes all messages that belong to a different experiment; 
in this case, we are only interested in blue marbles.

\item The \texttt{max\_iterations} parameter indicates the number of populations that
are going to be accepted. When this number of messages is reached, we must
end the loop. In this example, \texttt{max\_iterations = 6},  the take operator 
assures that only 6 messages are received. After the Observable emits 
the sixth message, it triggers the completed event.

\item Many observers can be subscribed to the \texttt{valid\_messages}  Observable because it
emits all the valid items. At the moment, there are to additional methods
subscribed that we are not showing, one for logging and the other for monitoring
the search.

\item The   \texttt{buffer\_with\_count(3)} operator, waits until it receives 
three valid messages to emit a single message that contains a list with
the previous three. The migration method requires that list to mix them. 
In our previous work, we needed a local buffer for storing a certain 
number of populations to mix them with others. This design has the advantage
of not needing extra memory, and it integrates better with the reactive paradigm.
A possible disadvantage can be that it only mixes contiguous populations,
but this can be mitigated with a larger buffer.

\item The  \texttt{population\_mixer} method pushes new populations to the messages Observable,
a  \texttt{publish(population)} observer is responsible for pushing 
the newly generated populations back to the Input Message Queue.
\end{enumerate}

@@

Figure \ref{fig:worker}.

\begin{algorithm}
    \caption{Migration}
    \label{read}
    \begin{algorithmic}[1]
        \Procedure{cxBestFromEach}{$pop_1,pop_2$}
            \State $pop_1.sort()$
            \State $pop_2.sort()$
            \State $size\gets min(len(pop_1), len(pop_2))$
            \State $cxpoint\gets (size-1)/2$
            \State $pop_1[cxpoint:]\gets pop_2[:cxpoint+2]$
            \State \textbf{return} $pop_1$
        \EndProcedure

    \end{algorithmic}
\end{algorithm}

\subsection{Experimental Process} 
\label{experiment_flow}

@@

>>>>

\subsection{Event Driven Model} 
\label{edm}
Based on a reactive architecture we proposed in a previous work
\cite{guervos2018introducing}, we now describe the general architecture shown in
Figure \ref{fig:kafkEO}. Again, we can explain the model using the analogy
of producers and consumers of messages. First, we can see that
there are two queues, one labeled \texttt{Input}, and the other one \texttt{Output}. In the diagram, 
push operations on a queue are represented by solid arrows connecting to the left side
of the queue box, and pull or pop operations as solid arrows leaving from the right side.
Also, the architecture has at least four processes indicated in the diagram as
swimlanes: First, the \texttt{Setup} process, responsible for the reading a configuration
file, and creating the initial subpopulations. Second, the \texttt{Controller} process,
responsible for the migration between subpopulations, and keeping track of the
iterations of the algorithm. The \texttt{Message Queue} process runs the \texttt{Input} and \texttt{Output}
queues. Finally, there is at least one \texttt{Stateless Function} process responsible
for running the isolated algorithms. In the example, two processes \texttt{PSO} and \texttt{GA},
are shown.

The algorithm starts with the \texttt{Setup} process pulling a configuration message from
the \texttt{Experiments} queue, this is not shown in the diagram. The configuration message 
includes all the parameters needed to execute the algorithm, among other things,
the number of subpopulations, the number of individuals, and the number of iterations
of the algorithm. We give more details about the configuration data structure
in Section \ref{experiment_flow}. % Missing  % added 
Once the configuration is read, we can follow the path of 
messagess as follows:

\begin{enumerate}
\item In this step, the specified number of populations are created according to the 
parameters found in the configuration structure. The population at this moment is just
static data, including each individual inside. Each subpopulation includes a metadata
section where its algorithm and execution parameters are specified. For instance, 
for a GA, the mutation rate, type of crossover operator, and other values are indicated.

\item Each subpopulation is then pushed to the \texttt{Input} queue, so they can be consumed 
by Stateless functions responsible for the execution of the search.  % Why the caps in Stateless? - JJ  
% It refers to the swimlines in the diagram, maybe I should point it out  - Mario
% Added a comment above. - Mario

\item One or more \texttt{Stateless functions} are constantly pulling subpopulation messages
from the \texttt{Input} queue. Taking the population data and metadata as parameters, these 
functions are responsible for the actual execution of the optimization algorithm. 
They take the current state of the population and run the algorithm for a certain 
number of iterations.

@@

\item The \texttt{Controller} process is responsible for keeping track of the progress of 
the search. It pulls current populations from the \texttt{Output} queue, inspects the metadata 
and if an optimal solution has been found or the maximum number of iteration has been
reached it signals the stop of the execution.

\item Otherwise, it passes the stream of messages 
to a migration process, where subpopulations are mixed with each other. New populations 
are generated from this migration, and they are again pushed to the Input queue to continue 
in a loop. The \texttt{Controller} is also responsible of logging the metadata recived with the
messages.  
\end{enumerate}

@@

On the other hand, there are several caveats as well:
It is more costly to move entire populations as messages than only passing certain individuals 
from a process to another. Designers must consider this cost when working with large
individuals, and if possible send populations using several messages, use compression or 
adjust the size of populations. Pool-based algorithms also suffer from this drawback, but 
web based implementations have been working with continuous optimization problems of 1000 dimensions.
But, in other cases this is not a viable approach.

@@

Next, we compare the message-driven architecture we presented against four other
proposals for multi-population algorithms found in the literature. We center the
comparison on the coupling and communication between the main components:
subpopulations, processing nodes, and algorithms.  In Figure \ref{fig:population_message} we show the
main components of a message-driven architecture. We can see that the algorithm
and subpopulations are separated, and the \texttt{Population queue} keeps the state
of subpopulations. While algorithms only need the populations as parameters
for their execution.

@@

In this section, we propose a design of a reactive container-based application
for executing multi-population based optimization experiments.  For the design,
we followed the design patterns highlighted in the previous section, and again
we can go back to Figure \ref{fig:kafkEO} and use it as a guide for explaining
the main components and their interactions.

% But first, we give a brief description of the technology we used.

% Add if needed
%\subsection{Containers} 
%\label{containers}
% TO DO

%\subsection{Docker Compose} 
%\label{compose}
% TO DO

% Do we need a description of every container? Wouldn't a general 
% description  of the containers used be enough? - JJ 
% Ok, I thin only workers and the controller need to be described in detail - Mario 
\subsection{Message Queue Container} 
\label{message_container}

In this work, we implemented all message
queues in the Redis memory store. Each queue is a Redis List object, and we use
the \texttt{LPOP} (left pop), and \texttt{RPOP} (right pop) commands for a queue like behavior.
In those cases where we needed a blocking behavior, i.e., when a process needs
to wait until a message is available, we used the blocking versions \texttt{BLPOP} and
\texttt{BRPOP}. Redis in-memory operation are very fast, and all of the obove 
operations have a time complexity of $\mathcal{O}(1)$. We use the official 
\texttt{redis:alpine} container image from DockerHub.

@@

The controller is an essential component of the architecture (see Figure \ref{fig:kafkEO})
because it is responsible for maintaining the evolutionary loop. It takes newly
evolved populations from the \texttt{Output Queue}, mixes them,   and produces new
messages from the result. At the same time, it must keep track of two conditions
for ending the loop: the number of messages it has received reaches a maximum
value, or the error of the best solution goes below a specific threshold.  Finally, it
must filter out remnant messages from other experiments. Messages from other
experiments can remain in the queues because of the asynchronous nature of the
system.  We chose to implement the controller in Python but using an API
specialized in asynchronous event-driven programming over data streams.  
The open-source library is called Reactive Extensions (ReactiveX) for Python 
\url{https://github.com/ReactiveX/RxPY} an it is based on the Observer 
pattern \cite{gamma1995design} and functional programming.

@@

In ReactiveX, an \texttt{observer} object subscribes to an \texttt{Observable} instance. 
Observables emit a sequence of items, and all the subscribed observers react to each emission. The
ReactiveX library includes several reactive operators that we can use to
transform and combine sequences of items. These operators provide reactive
extensions that allow us to compose asynchronous sequences together in a
declarative manner. In Figure \ref{fig:marble_controller}, we use a marble diagram to
represent the composition of \texttt{Observables} and reactive extensions operators. We
represent the timeline of an \texttt{Observable} as a horizontal arrow, which indicates
that time flows from left to right. In the diagram, items are represented as
marbles. The position of each marble indicates the point in time when they were
emitted by the \texttt{Observable}. Reactive operators are represented as text boxes,
showing the transformation to be applied. Normally, a transformation results in
another \texttt{Observable}, again emitting new results.  For asynchronous programming, 
this approach is more elegant than nested callbacks, that are more difficult 
to code and debug.

@@

\begin{enumerate}
\item The controller continually pulls new messages from the \texttt{Output} 
message queue. These messages are instantly emitted by the  \texttt{consumed\_messages}
\texttt{Observable}. In this particular period, we can see that the stream receives 
two populations from another experiment, and are shown as two red marbles.  
The filter operator removes all messages that belong to a different experiment; 
in this case, we are only interested in blue marbles.

\item The \texttt{max\_iterations} parameter indicates the number of populations that
are going to be accepted. When this number of messages is reached, we must
end the loop. In this example, \texttt{max\_iterations = 6},  the take operator 
assures that only 6 messages are received. After the \texttt{Observable} emits 
the sixth message, it triggers the completed event.

\item Many observers can be subscribed to the \texttt{valid\_messages} \texttt{Observable} because it
emits all the valid items. At the moment, there are to additional methods
subscribed that we are not showing, one for logging and the other for monitoring
the search.

\item The   \texttt{buffer\_with\_count(3)} operator, waits until it receives 
three valid messages to emit a single message that contains a list with
the previous three. The \texttt{population\_mixer}  method requires that list to mix them. 
In our previous work, we needed a local buffer for storing a certain 
number of populations to mix them with others. This design has the advantage
of not needing extra memory, and it integrates better with the reactive paradigm.
A possible disadvantage can be that it only mixes contiguous populations,
but this can be mitigated with a larger buffer.

\item The \texttt{population\_mixer} receives a list of three populations,
let us say [A, B, C], and calls the migration method shown in
Algorithm \ref{alg:migration}  for [A, B], [B, C] and [A, C]. 
The migration algorithm sorts both populations and generates
a new population containing the best half from each.
This migration method is similar to those used successfully in previous 
works on pool-based algorithms \ref{fig:pool_island}.  
Finally, the method pushes the new populations to the 
\texttt{messages Observable}. From there, a
\texttt{publish(population)} observer is responsible for pushing
the newly generated populations back to the \texttt{Input} message queue.
\end{enumerate}

\begin{algorithm}
    \caption{Migration}
    \label{alg:migration}
    \begin{algorithmic}[1]
        \Procedure{cxBestFromEach}{$pop_1,pop_2$}
            \State $pop_1.sort()$
            \State $pop_2.sort()$
            \State $size\gets min(len(pop_1), len(pop_2))$
            \State $cxpoint\gets (size-1)/2$
            \State $pop_1[cxpoint:]\gets pop_2[:cxpoint+2]$
            \State \textbf{return} $pop_1$
        \EndProcedure 
    \end{algorithmic}
\end{algorithm}

In the next section, we follow the algorithm flow and describe the 
implementation details of the \texttt{worker} containers.

@@

Worker containers include a Python script called \texttt{main.py} running an
infinite loop, continually trying to pull a new population to work on it. Once
\texttt{main.py} receives a message containing configuration data, it creates a
worker object responsible for initializing and running the specified stateless
function (i.e., GA or PSO). Once the function returns a new population, the
worker pushes the results to the \texttt{Output} queue, finishes, and passes
the execution to the main loop. This process is shown in Figure
\ref{fig:worker}.

\subsection{Experimental Workflow in EvoSwarm} 
\label{experiment_flow}

@@

We aimed to learn if the proposed solution can be an efficient solution to
improve the scalability and performance of population-based optimization
algorithms. Hence, in this section, we verify the following:

\begin{itemize}
 \item If we can improve the execution time of the algorithm by adding populations
 and serverless functions to the system. In particular, what is the effect of
 these changes on the scalability of the system?

\item Having a multi-population enabled platform, with heterogeneous populations and
the support for mixing search strategies, can increase the performance of the
search by needing fewer function evaluations than a homogeneous setting?

\end{itemize}

To validate these questions, we used benchmark functions from the Continuous
Noiseless BBOB testbed, which is part of the Comparing Continuous Optimizers
(COCO) framework \cite{hansen2016coco}. The testbed includes 24 real-parameter,
single-objective benchmark functions, and the capability to provide additional
instances of each function. Each instance of a function has a different optimal
value. The standard benchmark of the testbed utilizes 15 instances per function
over 2, 3, 5, 10, 20, and 40 dimensions. The maximum number of Function
Evaluations (FEs) changes according to the dimension (D), and it is determined
by the expression $10^5 \cdot D$. As an example, if we have $D = 2$, the
maximum number of FEs is $200,000$

The COCO framework offers several tools to compare the performance of
algorithms, generating data sets, tables, and reports for an experiment. There
is a repository of more than 200 results for the noiseless BOBB testbed,
collected from BBOB workshops and special sessions between the years 2009 and
2019 \url{https://coco.gforge.inria.fr/doku.php?id=algorithms-bbob}. The
EvoSwarm application includes an adapted version of the noiseless BOBB testbed,
compatible with the scripts of the framework, to compare with other algorithms
in the repository.

To validate the heterogeneous multi-population setting, we compare between
homogeneous and an ensemble of multi-populations, using Genetic Algorithms
(GAs) and Particle Swarm Optimization (PSO). For the GA implementation, we used
the DEAP library \cite{fortin2012deap} and for the PSO, the EvoloPy library
\cite{faris2016evolopy}. Both Python libraries are open-source. In EvoSwarm 
these two algorithms are implemented as stateless functions.

We show the parameters for each algorithm, GA in Table \ref{tab:GAparams} and
for PSO in Table \ref{tab:PSOparams}. We obtained these parameters in a
previous work \cite{}. To obtain the parameters, we tested first on the
Rastrigin separable function with five dimensions. After about fifteen
experiments, the most challenging targets were achieved for this particular
function. We tested again with functions one to three, and after obtaining
favorable results, the PSO and GA algorithm parameters were set.In the GA, we
randomly set the mutation and crossover probabilities to have more
heterogeneous workers; in these parameters, the range of values is specified.

\begin{table}
  \small
  \caption{ DEAP GA EvoWorker Parameters }
  \label{tab:GAparams} 
  \centering
  \small
  \begin{tabular}{|l|c|}
    \hline
    Selection & Tournament size=12\\ \hline
    Mutation & Gaussian $\mu=0.0$, $\sigma=0.5$, indbp=0.05  \\ \hline
    Mutation Probability & [0.1,0.3]  \\ \hline
    Crossover & Two Point  \\ \hline
    Crossover Probability& [0.2,0.6]  \\ \hline
  \end{tabular}
\end{table}

\begin{table}
  \small
  \caption{ EvoloPy PSO Parameters }
  \label{tab:PSOparams} 
  \centering
  \small
  \begin{tabular}{|l|c|}
    \hline
    $V_{max}$ & 6 \\ \hline
    $W_{max}$ & $0.9$ \\ \hline
    $W_{min}$ & $0.2$ \\ \hline
    $C_1$ & 2 \\ \hline
    $C_2$ & 2 \\ \hline
  \end{tabular}
\end{table}

\begin{table}
    \small
    \caption{10 Populations experiment parameters
    }
    \label{tab:params:workers}
    \centering
    \small
    \begin{tabular}{|l|c|c|c|c|c|c|}
      \hline
      Dimension        & 2  & 3  & 5  & 10 & 20  & 40  \\ \hline
      Generations      & 40 & 25 & 28 & 50 & 66  & 80  \\ \hline
      Population Size  & 50 & 60 & 60 & 70 & 100 & 125 \\ \hline
      Populations      & 10 & 10 & 10 & 10 & 10  & 10  \\ \hline
      Iterations       & 10 & 20 & 30 & 30 & 30  & 40  \\ \hline  
    \end{tabular}
\end{table}

\begin{table}
    \small
    \caption{5 Populations experiment parameters
    }
    \label{tab:params:workers}
    \centering
    \small
    \begin{tabular}{|l|c|c|c|c|c|c|}
      \hline
      Dimension        & 2  & 3  & 5  & 10 & 20  & 40  \\ \hline
      Generations      & 40 & 25 & 28 & 50 & 66  & 80  \\ \hline
      Population Size  & 100 & 120 & 120 & 140 & 200 & 250 \\ \hline
      Populations      & 5 & 5 & 5 & 5 & 5  & 5  \\ \hline
      Iterations       & 10 & 20 & 30 & 30 & 30  & 40  \\ \hline  
    \end{tabular}
\end{table}
